---
title: "Life Expectancy Over Time by Continent"
output:
  html_document: default
date: "2024-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Background
Life expectancy is a key indicator of population health and socio-economic development, reflecting the overall quality of life in a country. This analysis examines global trends in life expectancy from 1990 to 2022, focusing on differences across continents.

## Data Origins
The data was obtained from the [World Bank Group](https://databank.worldbank.org/source/world-development-indicators), where I manually selected the variables I was interested in. Under database, I selected life expectancy at birth, females, life expectancy at birth, males, life expectancy at birth, total. Under country, I manually selected all 196 countries, and for time, I selected the years 1990 to 2023. In total, the data has 589 data entries.

## Research Question
**How has life expectancy across different continents change throughout the years?**
It could be interesting to look at changes in average life expectancy as it reflects the health, development, and well-being of populations. Examining how it has changed across continents over time highlights global progress and disparities in healthcare, economic growth, and social policies. Gaining insights into these trends can help to promote health equity and sustainable development by highlighting areas that require attention.

```{r, include=FALSE}
#Loading necessary packages
library(readxl)
library(here)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(countrycode)
library(gganimate)
library(gifski)
```

```{r}
#Loading data into R
data <- read_excel(here("data", "WorldLifeExp.xlsx"))
#Displaying the data
head(data)
```

## Data Cleaning
```{r}
#Removing unneccessary columns
data <- data %>%
  select(-`Country Code`, -`Series Code`, -`2023 [YR2023]`)  

#Removing rows with no value assigned
data <- data %>%
  filter(!if_any(everything(), ~ . == ".."))

#Renaming columns so it is easier to read
data <- data %>%
  rename(`1990` = `1990 [YR1990]`, `1991` = `1991 [YR1991]`, `1992` = `1992 [YR1992]`, `1993` = `1993 [YR1993]`, `1994` = `1994 [YR1994]`, `1995` = `1995 [YR1995]`, `1996` = `1996 [YR1996]`, `1997` = `1997 [YR1997]`, `1998` = `1998 [YR1998]`, `1999` = `1999 [YR1999]`, `2000` = `2000 [YR2000]`, `2001` = `2001 [YR2001]`, `2002` = `2002 [YR2002]`, `2003` = `2003 [YR2003]`, `2004` = `2004 [YR2004]`, `2005` = `2005 [YR2005]`, `2006` = `2006 [YR2006]`, `2007` = `2007 [YR2007]`, `2008` = `2008 [YR2008]`, `2009` = `2009 [YR2009]`, `2010` = `2010 [YR2010]`, `2011` = `2011 [YR2011]`, `2012` = `2012 [YR2012]`, `2013` = `2013 [YR2013]`, `2014` = `2014 [YR2014]`, `2015` = `2015 [YR2015]`, `2016` = `2016 [YR2016]`, `2017` = `2017 [YR2017]`, `2018` = `2018 [YR2018]`, `2019` = `2019 [YR2019]`, `2020` = `2020 [YR2020]`, `2021` = `2021 [YR2021]`, `2022` = `2022 [YR2022]`)
data <- data %>%
  rename(`Population Group` = `Series Name`)

#Converting relevant columns to numeric
data <- data %>%
  mutate(across(c(`1990`, `1991`, `1992`, `1993`, `1994`, `1995`, `1996`, `1997`, 
                  `1998`, `1999`, `2000`, `2001`, `2002`, `2003`, `2004`, `2005`, 
                  `2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, 
                  `2014`, `2015`, `2016`, `2017`, `2018`, `2019`, `2020`, `2021`, `2022`), 
                ~ as.numeric(as.character(.))))

#Rounding up life expectancy values to two decimal points
data <- data %>%
  mutate(across(where(is.numeric), ~ round(., 2)))

#Renaming variables so it is easier to read 
data <- data %>%
  mutate(`Population Group` = recode(
    `Population Group`,
    `Life expectancy at birth, total (years)` = "Total",
    `Life expectancy at birth, male (years)` = "Male",
    `Life expectancy at birth, female (years)` = "Female"
  ))

```

## Data Preparation
As there are so many countries, it would be difficult to visualise the data efficiently, hence, I used the countrycode package to sort the countries into their specific continents
```{r warning = FALSE}
#Correcting the way some countries were typed so it matches data in countrycode package
manual_corrections <- data.frame(
  Country = c("Bahamas, The", "Congo, Rep.", "Congo, Dem. Rep.", "Egypt, Arab Rep.", "Gambia, The.", "Hong Kong SAR, China", "Korea, Rep.", "Iran, Islamic Rep.", "Korea, Dem. People's Rep.", "Lao PDR", "Kyrgyz Republic", "Micronesia, Fed. Sts.", "Russian Federation", "Syrian Arab Republic", "Venezuela, RB", "Yemen, Rep.", "Viet Nam"),
  Corrected = c("Bahamas", "Republic of the Congo", "Democratic Republic of the Congo", "Egypt", "Gambia", "Hong Kong", "South Korea", "Iran", "North Korea", "Laos", "Kyrgyzstan", "Micronesia (Federated States of)", "Russia", "Syria", "Venezuela", "Yemen", "Vietnam")
)

data <- data %>%
  left_join(manual_corrections, by = c("Country Name" = "Country")) %>%
  mutate(`Country Name` = if_else(is.na(Corrected), `Country Name`, Corrected)) %>%
  select(-Corrected)

#Using the countrycode package and creating a new column for continent and sorting countries under it
data <- data %>%
  mutate(Continent = countrycode(`Country Name`, origin = "country.name", destination = "continent"))

#Manually assigning Kosovo to Europe as countrycode couldn't identify its continent
data <- data %>%
  mutate(Continent = if_else(`Country Name` == "Kosovo", "Europe", Continent))

# Renaming "Americas" to "North & South America" in the Continent column
data <- data %>%
  mutate(Continent = if_else(Continent == "Americas", "North & South America", Continent))

#Moving the Continent column next to the Country Name column for better readability
data <- data %>%
  select(1, Continent, everything())
```

I filtered out the total life expectancy rows so I can compare total life expectancy across continents
```{r}
total_data <- data %>%
  filter(`Population Group` == "Total")
```

Next, I converted the data frame into a long format
```{r}
#Reshaping my data into long format 
long_data <- total_data %>%
  pivot_longer(
    cols = `1990`:`2022`,  # Columns for each year
    names_to = "Year",     # Name for the new column with years
    values_to = "Life Expectancy"  # Name for the new column with values
  ) %>%
  mutate(Year = as.numeric(Year))  # Convert year to numeric
```

Finally, I calculated the average life expectancy for each continent and year
```{r, echo=TRUE, results="hide"}
continent_data <- long_data %>%
  group_by(Continent, Year) %>%
  summarise(Average_Life_Expectancy = mean(`Life Expectancy`, na.rm = TRUE), .groups = "drop")

str(continent_data$Year) #Ensuring year column is properly structured
range(continent_data$Year, na.rm = TRUE) #Check range

continent_data <- continent_data %>% filter(!is.na(Year) & Year != Inf) #Removes rows with invalid or missing year values
```

## Initial visualisation
This is the first graph that I plotted with the data above

```{r}
p <- ggplot(continent_data, aes(x = Year, y = Average_Life_Expectancy, color = Continent, group = Continent)) +
  geom_line(linewidth = 1.2, alpha = 0.7) +  #Lines for each continent
  labs(title = "Life Expectancy Over Time by Continent",
    x = "Year",
    y = "Average Life Expectancy") 

print(p)

#Saving the graph into figs folder
ggsave("figs/initialgraph.png", plot = p, width = 8, height = 6, dpi = 300)
```
But this is a bad graph as the year and average life expectancy values are spread too far out, making it hard to visualise the graph. 

## Improving on initial visualisation
Therefore, I improved my code and added these few things:
1. Changed the font size of my graph and centralised the title and subtitle 
2. Changed the colour of the continents according to the Olympics colour coding system for each continent
3. Coded every year to appear on the graph for better visualisation
4. Adjusted the y-axis range so that average life expectancy values are spaced by intervals of 5 

```{r}
#Tidying the graph and making it look better
graph_theme <- theme(
  plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),  #Main title in bold and centralised
  plot.subtitle = element_text(hjust = 0.5, size = 12, face = "italic"),  #Subtitle in italic and centralised
  axis.text.x = element_text(angle = 45, hjust = 1),  #Rotate x-axis labels
  legend.title = element_text(face = "bold", size = 12),  #Legend title in bold
  legend.text = element_text(size = 12)  #Legend text size
)

#Colour coding each continent according to the Olympics
continent_colors <- c(
  "Europe" = "#0066B3",   #Blue
  "Asia" = "#FFCC00",     #Yellow
  "Africa" = "#000000",   #Black
  "Oceania" = "#009639",  #Green
  "North & South America" = "#F44336"  #Red
)
```

```{r, fig.width = 12, fig.height = 6}
#Plotting a better graph with the changes
p2 <- ggplot(continent_data, aes(x = Year, y = Average_Life_Expectancy, color = Continent, group = Continent)) +
  geom_line() +  
  geom_point(size = 2) +  # Points for each year for each continent
  labs(
    title = "Life Expectancy Over Time by Continent",
    subtitle = "Year: 1990 to 2022",
    x = "Year",
    y = "Average Life Expectancy") +
  scale_x_continuous(
    breaks = seq(min(continent_data$Year), max(continent_data$Year), by = 1)) + #Show a label every year
  scale_y_continuous(
    breaks = seq(0, 100, by = 5),  #Adjust y-axis ticks for more specific ranges
    limits = c(0, 100) ) +            #Set the y-axis range (adjust as needed)
  scale_color_manual(values = continent_colors) +  #Apply Olympic colours
  graph_theme

print(p2)

#Saving the plot
ggsave("figs/secondgraph.png", plot = p2, width = 12, height = 6, dpi = 300)
```

